Synthesising Complex Termination Arguments

### 2014-06-18 papers/termination/paper.tex
minor revision

Bit-vector arithmetic is very similar to modular arithmetic.  For a computer using $k$-bit words, unsigned
calulcations are just defined to 
===============
The abstraction of bit-vectors to mathematical integers 
ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in 
incomparable behaviours with respect to termination. 
%The loop in Figure~\ref{} terminates for integers, but does not terminate for bit-vectors. 
The loop in Figure~\ref{} always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
%% =======
%% most of the work concentrated on finding \emph{linear ranking functions for linear programs over integers}.
%% Next, we will consider in more detail the assumptions made by such techniques and their implications.\\

@@

%% The abstraction of bit-vectors to mathematical integers 
%% ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in %, it gives rise to 
%% %{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors are ubiquitous in computer systems. 
%% incomparable behaviours with respect to termination:
%% %The differences between the semantics of unbounded mathematical integers and that of machine integers result in 
%% \begin{itemize}
%% \item Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
%% \begin{lstlisting}[language=C]
%% while (x > 0) x -= 2;
%% \end{lstlisting}
%% The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
%% >>>>>>> 7b690c08a70e4161a432600e9456d922106badaf
whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
According to the C11 standard for the C programming language, a signed under- or over-flow yields undefined behaviour.
However, compilers generally treat signed under- and over-flows using the
wrap-around behaviour (see Section~\ref{sec:machine.arith}). %This means that a termination analysis for integers would return an unsound result for this situation.
As a second example, the loop in Figure~\ref{} is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.

>>>>

%Bit-vector arithmetic is very similar to modular arithmetic.  For a computer using $k$-bit words, unsigned
%calulcations are just defined to 
The abstraction of bit-vectors to mathematical integers 
ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in 
incomparable behaviours with respect to termination. 
%The loop in Figure~\ref{} terminates for integers, but does not terminate for bit-vectors. 
The loop in Figure~\ref{} always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
%% =======
%% most of the work concentrated on finding \emph{linear ranking functions for linear programs over integers}.
%% Next, we will consider in more detail the assumptions made by such techniques and their implications.\\

@@

%% The abstraction of bit-vectors to mathematical integers 
%% ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in %, it gives rise to 
%% %{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors are ubiquitous in computer systems. 
%% incomparable behaviours with respect to termination:
%% %The differences between the semantics of unbounded mathematical integers and that of machine integers result in 
%% \begin{itemize}
%% \item Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
%% \begin{lstlisting}[language=C]
%% while (x > 0) x -= 2;
%% \end{lstlisting}
%% The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
%% >>>>>>> 7b690c08a70e4161a432600e9456d922106badaf
According to the C11 standard for the C programming language, a signed under- or over-flow yields undefined behaviour.
However, compilers generally treat signed under- and over-flows using the
wrap-around behaviour (see Section~\ref{sec:machine.arith}). %This means that a termination analysis for integers would return an unsound result for this situation.
As a second example, the loop in Figure~\ref{} is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.
